A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship. Robert Atkey is supported by EPSRC grant FIXME.}

### 2022-01-26 tex/esop22/env.tex
[ edit ] various additions and changes

Given a semantic notion of variable $\V$ : \AgdaFunction{OpenFam}, we
use the notation $\Gamma \sdtstile{}\V A$ meaning $\V\,\Gamma\,A$ for
the type of inhabitants of $\V$ in the context $\Gamma$ at type
$A$. In the non-substructural systems of Allais et al. \cite{AACMM21},
a $\V$-environment $\Gamma \env\V \Delta$ is nothing more than a
function $\forall A \to \Delta \sqni A \to \Gamma \sdtstile{}\V A$,
mapping variables to $\V$-things. In our usage annotated setting
though, we must correctly distribute resources tracked by the
annotations; making sure that we have enough resources in $\Gamma$ to
cover all the demands in $\Delta$. Following the work of \citet{WA20},
this accounting is expressed via the presence of a linear
transformation:

@@

\begin{definition}[Environment]\label{def:lr-env}
  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and
  $\Delta$ (decomposed as $\grP\gamma$ and $\grQ\delta$, respectively,
  when convenient) is a linear map
  $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$,
  $\grPprime$, and $\grQprime$ such that
  $\grPprime \leq \grQprime\gr\Psi$, a function from
  $\grQprime\delta \sqni A$ to $\grPprime\gamma \sdtstile{}\V A$.

  \bob{Write out the Agda notation for environments}
\end{definition}

By instantiating $\V$, we obtain resource correct versions of familiar
notions: letting $\V$ be $\sqni$ yields resource correct renamings;
and letting $\V$ be $\vdash$ (i.e., terms) yields resource correct
substitutions.

@@

As a mnemonic, one may use notation like the following to see what values are
needed in the environment.
\[
  \begin{pmatrix}
    \gr0A & \gr0\plr{B \multimap C} & \gr1B \\
    \gr0A & \gr1\plr{B \multimap C} & \gr1B
  \end{pmatrix}
  \begin{matrix}
    {} \vdash B \\
    {} \vdash C
  \end{matrix}
\]
This notation assumes that $\V$ supports subusaging, which is
always the case when we are using environments for traversals.\bob{Forward ref to where this is explicitly assumed?}

>>>>

Given a semantic notion of variable $\V$ : \AgdaFunction{OpenFam}, we
use the notation $\Gamma \sdtstile{}\V A$ meaning $\V\,\Gamma\,A$ for
the type of inhabitants of $\V$ in the context $\Gamma$ at type
$A$. In the non-substructural systems of Allais et al. \cite{AACMM21},
a $\V$-environment $\Gamma \env\V \Delta$ is nothing more than a
function $\forall A \to \Delta \sqni A \to \Gamma \sdtstile{}\V A$,
mapping variables to $\V$-things. In our usage annotated setting
though, we must correctly distribute resources tracked by the
annotations; making sure that we have enough resources in $\Gamma$ to
cover all the demands in $\Delta$. Following our previous work~\cite{WA20},
this accounting is expressed via the presence of a linear
transformation:

@@

\begin{definition}[Environment]\label{def:lr-env}
  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and
  $\Delta$ (decomposed as $\grP\gamma$ and $\grQ\delta$, respectively,
  when convenient) is a linear map
  $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$,
  $\grPprime$, and $\grQprime$ such that
  $\grPprime \leq \grQprime\gr\Psi$, a ``lookup'' function from
  $\grQprime\delta \sqni A$ to $\grPprime\gamma \sdtstile{}\V A$.
\end{definition}

In Agda code, we use \ExecuteMetaData[\Snippetstex]{GvDashA}\ instead of
$\Gamma \sdtstile{}\V A$ and \ExecuteMetaData[\Snippetstex]{GenvD}\ instead of
$\Gamma \env\V \Delta$.

The specification of the lookup function has some redundancy.
Notice that, for $\grQprime\Delta \sqni A$ to hold, we must have
$\grQprime \leq \langle i \rvert$ for some $i$.
Instead of $\grPprime \leq \grQprime\gr\Psi$, asking for
$\grPprime \leq \langle i \rvert\gr\Psi$ would be just as general.
Additionally, all of the $\V$s we consider satisfy the \emph{subusaging}
property (that $\grPprime \leq \grP$ yields a coercion
$\grP\Gamma \sdtstile{}\V A \to \grPprime\Gamma \sdtstile{}\V A$), in which case
we could just ask for an inhabitant of
$\plr{\langle i \rvert\gr\Psi}\gamma \sdtstile{}\V A$.
However, we find the stated definition technically expedient because, by this
point, basis vectors and raw indices (instead of usage-checked variables) are
below our level of abstraction.
We prefer to work with linear relatedness and $\sqni$-variables.

By instantiating $\V$ in \cref{def:lr-env}, we obtain resource-correct versions
of familiar
notions: letting $\V$ be $\sqni$ yields resource-correct renamings;
and letting $\V$ be $\vdash$ (i.e., terms) yields resource-correct
substitutions.

@@

As a mnemonic, one may use notation like the following to see what values are
needed in the environment.
\[
  \begin{pmatrix}
    \gr0A & \gr0\plr{B \multimap C} & \gr1B \\
    \gr0A & \gr1\plr{B \multimap C} & \gr1B
  \end{pmatrix}
  \begin{matrix}
    {} \vdash B \\
    {} \vdash C
  \end{matrix}
\]
This notation assumes that $\V$ supports subusaging.

### 2022-01-26 tex/esop22/quant-framework.tex
[ esop22 ] Reorder Section 6, try to smooth narrative
[ edit ] various additions and changes

\newcommand{\todo}[1]{}

@@

>>>>

\newcommand{\todo}[1]{\textbf{#1}}

@@

\appendix

{\bf TODO: explain and format properly.}
\AgdaInductiveConstructor{!E$'$} and \AgdaInductiveConstructor{!I$'$} are
variants on \AgdaInductiveConstructor{!E} and \AgdaInductiveConstructor{!I}
with some of the usage contexts given explicitly (as well as the proofs that the
usage contexts are correctly related).
On concrete posemirings, unification is particularly poor at inferring these
usage contexts from the proofs because addition and multiplication are no longer
(judgementally) injective.
\ExecuteMetaData[\HeavyItex]{cojoin-explicit}

### 2022-01-26 tex/esop22/semantics.tex
[ edit ] various additions and changes

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

>>>>

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.
